---
title: |
  <div class="tight-title">
    <div>ECOM30004/90004</div>
    <div>Time Series Analysis and Forecasting</div>
     <div class="spacer"></div> <!-- blank line -->
    <div class="bold">Tutorial 9</div>

    <div class="spacer"></div> <!-- blank line -->

    <div class="author">Tutor: Pengyun (Kevin) Chen</div>
    <div class="author">University of Melbourne</div>
    <div class="author">Semester 2, 2026</div>
  </div>

  
format:
  revealjs:
    theme: simple
    self-contained: true
    chalkboard: false
    width: 1300
    height: 950
    footer: 'Access Slides: `pengyunc.github.io/TSAF_Week9`'
    title-slide-attributes:
      data-background-color: "#000F46"
    slide-number: true
    center-title-slide: true
    toc: false
    css: styles.css
    callout-appearance: simple
    background-transition: fade
    callout-icon: true
    transition: slide
    scrollable: true
    code-block-height: 800px
    code-line-numbers: true
    mathjax: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-svg.js"
    html-math-method: mathjax
    execute:
      echo: true
      eval: true
      enabled: true
---

## Notice

::: {.callout-warning title="Conditions of Use"}

- This document is intended for personal use and is NOT meant to be shared on the internet (e.g. Studocu, ThinkSwap, StudentVIP, Coursehero, etc.)
- Do not share this document with anyone, even if they are enrolled in ECOM30004/90004.
- Official course materials, such as lecture slides, tutorial sheets and prescribed readings should always be your primary point of reference.

:::
<div style="height: 0.2em;"></div>
 - Slides accessible here: `pengyunc.github.io/TSAF_Week9`
 - Download HTML versions here: [Dropbox Link](https://www.dropbox.com/scl/fo/aoxoud5ex0j42hsrod3dg/ADblAt_jJ9hpESZME9FDcWc?rlkey=0s0oyf1y3uokzvlhbzxe2yy3g&st=kx2yupzr&dl=0)




## This tutorial
<blue>This tutorial uses lag polynomials to (A) understand and (B) extend our ARMA modelling approach.</blue>
<div style="height: 0.5em;"></div>
:::{.callout-note title="Agenda"}
- We will make use of `BAB3mth.csv` from Week 4 and `RetailSales.csv` from Week 1.
- All code is provided in the questions in order to focus on the concepts and results.
- <blue>Part A: </blue>Common factors in ARMA models (`BAB3mth.csv`)
- <blue>Part B: </blue>Multiplicative models: seasonal autoregression (`RetailSales.csv`)
:::
<div class="horizontal-line"></div>
<blue>**Initialisation** for Part A</blue>
```{r}
library(urca)
library(forecast)

dt <- read.csv("BAB3mth.csv")
Y <- ts(dt$BAB3, start=c(2010,1), end=c(2025,6), frequency=12)
DY <- diff(Y)
Y <- window(Y, end=c(2024,6))
DY <- window(DY, end=c(2024,6))
```



## Recap - What's going on? {auto-animate="true"}

:::{.fragment}
Estimating an AR model is computationally simple.

- you can simply compute this by OLS (by matrix algebra, $\hat{\beta}=(X^{\top}X)^{-1}(X^{\top}y)$)
:::

:::{.fragment}
Estimating a model with an MA component is much more computationally intensive.

- you need to rely on numerical methods (e.g. maximum likelihood)
:::

:::{.fragment}

:::{.callout-tip}
Suppose we want to estimate the most basic ARMA(1,1).
$$y_{t}=\phi_{1}y_{t-1}+U_{t}+\theta_{1}U_{t-1}$$
The maximum likelihood algorithm proceeds as follows:

:::{.incremental}
- First, guess $(\phi_{1},\theta_{1})$.
- Next, plug these values into the model, and compute the implied residuals $U_{t}$.
- Compute the likelihood function (how probable the observed data are given those parameters)
- Repeat the process until the likelihood function is maximised.
:::

:::

:::

:::{.fragment}
<div style="height: 0.3em;"></div>
However, with invertibility, the MA component can be approximated by AR models.
:::

## What's going on? {auto-animate="true"}

However, with invertibility, the MA component can be approximated by AR models.


:::{.fragment}
<div class="horizontal-line"></div>
<blue>**What is invertibility?**</blue>

<blue>MA is invertible $\to$</blue> can be re-written as an AR($\infty$) representation <teal>(with converging coefficients)</teal>

:::

:::{.fragment}
<div class="horizontal-line"></div>

<blue>**Why can AR models approximate the MA component?**</blue>

:::{.callout-tip title="Lecture Example: MA(1)"}

$$\begin{aligned}
Z_{t}=U_{t}+\theta_{1}U_{t-1}\;\;\xrightarrow{\textsf{can be written as}}\;\;& Z_{t}=\sum^{\infty}_{j=1}-(-\theta_{1})^{j}Z_{t-j}+U_{t}\\
\textsf{or, when there is finite horizon:}\;\;& Z_{t}=\sum^{t-1}_{j=1}-(-\theta_{1})^{j}Z_{t-j}+U_{t}-(-\theta_{1})^{t}U_{0}
\end{aligned}$$

:::
<div style="height: 0.2em;"></div>
:::

:::{.fragment}
If $|\theta_{1}|<1$, then $(-\theta_{1})^{t}$ shrinks to 0 as $t$ becomes larger <green>$\to$ ignoring certain higher-order AR terms can still yield a reasonable approximation.</green>

:::

## What makes an MA model invertible? {auto-animate="true"}

<blue>**Example:** In Tutorial 8, the optimal ARMA model for $\Delta Y_{t}$ is ARMA(1,2) with drift.</blue>
\begin{align*}
\Delta{}Y_{t}&=\beta_{0}+Z_{t}\\
Z_{t}&=\phi_{1}Z_{t-1}+U_{t}+\theta_{1}U_{t-1}+\theta_{2}U_{t-2}
\end{align*}

:::{.fragment}
Consider only the second equation. Moving all AR terms to the LHS gives:
$$Z_{t}-\phi_{1}Z_{t-1}=U_{t}+\theta_{1}U_{t-1}+\theta_{2}U_{t-2}$$
:::

:::{.fragment}
Next, replace the notation with lag operators $L$, where $L^{n}(x_{t})=x_{t-n}$:


$$Z_{t}-\phi_{1}L(Z_{t})=U_{t}+\theta_{1}L(U_{t})+\theta_{2}L^{2}(U_{t})$$
:::

:::{.fragment}
$$(1-\phi_{1}L)Z_{t}=\color{green}{\underbrace{(1+\theta_{1}L+\theta_{2}L^{2})}_{\substack{\textsf{the MA lag polynomial}\\\textsf{we are interested in}}}}\color{black}{U_{t}}$$
:::


:::{.fragment}
Compute the solution of the <green>lag polynomial for the MA component</green>:

\begin{align*}
1+\theta_{1}z+\theta_{2}z^{2}&=0\\
\color{teal}{\boxed{\textsf{\small{}Suppose the solutions are $\to$}}}\qquad \zeta_{1},\zeta_{2}&=\frac{-\theta_{1}\pm\sqrt{\theta_{1}^{2}-4\theta_{2}}}{2\theta_{2}}
\end{align*}

:::

## What makes an MA model invertible? {auto-animate="true"}

Compute the solution of the <green>lag polynomial for the MA component</green>:

\begin{align*}
1+\theta_{1}z+\theta_{2}z^{2}&=0\\
\color{teal}{\boxed{\textsf{\small{}Suppose the solutions are $\to$}}}\qquad \zeta_{1},\zeta_{2}&=\frac{-\theta_{1}\pm\sqrt{\theta_{1}^{2}-4\theta_{2}}}{2\theta_{2}}
\end{align*}

:::{.fragment}
This means that the lag polynomial can be rewritten as:

$$(\zeta_{1}-z)(\zeta_{2}-z)=0$$
:::



::: {.fragment}
$$\color{teal}{\zeta_{1}}\color{black}{(1-\zeta_{1}^{-1}z)\,\times\,} \color{teal}{\zeta_{2}}\color{black}{(1-\zeta_{2}^{-1}z)=0}$$
:::




::: {.fragment}
$$(1-\zeta_{1}^{-1}z)(1-\zeta_{2}^{-1}z)=0$$
:::

:::{.fragment}

<blue>
The MA component is **invertible** if:
$$|\zeta_{1}^{-1}|<1\quad \text{and}\quad |\zeta_{2}^{-1}|<1.$$
Graphically, the inverse roots of the MA lag polynomial should lie **within** the unit circle in the Real-Imaginary plane.</blue>
:::


## What makes an MA model invertible? {auto-animate="true"}

<blue>
The MA component is **invertible** if:
$$|\zeta_{1}^{-1}|<1\quad \text{and}\quad |\zeta_{2}^{-1}|<1.$$
Graphically, the inverse roots of the MA lag polynomial should lie **within** the unit circle in the Real-Imaginary plane.
</blue>

<div class="horizontal-line"></div>

:::{.fragment}
<purple>**Aside.**</purple> Consider an ARMA(2,2) model:
\begin{align*}
(1-\phi_{1}L-\phi_{2}L^{2})Z_{t}&=(1+\theta_{1}L+\theta_{2}L^{2})U_{t}\\
(1-\xi_{1}^{-1}L)(1-\xi_{2}^{-1}L)Z_{t}&=(1-\zeta_{1}^{-1}L)(1-\zeta_{2}^{-1}L)U_{t}
\end{align*}
where $\xi_{1}^{-1},\xi_{2}^{-1}$ are inverse roots of the AR lag polynomial:
$1-\phi_{1}z-\phi_{2}z^{2}=0$.
<div style="height: 0.5em;"></div>
:::
:::{.fragment}
<purple>The AR component is **stationary** if:
$$|\xi_{1}^{-1}|<1\quad\text{and}\quad |\xi_{2}^{-1}|<1$$
Graphically, the inverse roots of the AR lag polynomial should lie **within** the unit circle in the Real-Imaginary plane.
</purple>
:::


## What makes an MA model invertible? {auto-animate="true"}

<blue>
The MA component is **invertible** if:
$$|\zeta_{1}^{-1}|<1\quad \text{and}\quad |\zeta_{2}^{-1}|<1.$$
Graphically, the inverse roots of the MA lag polynomial should lie **within** the unit circle in the Real-Imaginary plane.
</blue>


:::{.panel-tabset}

## Parameter Estimates
```{r}
eq <- Arima(DY, order=c(1,0,2))       # ARMA(1,2)
print(coef(eq))
```

<div style="height: 1em;"></div>

Thus the lag polynomial for the MA component is:
\begin{align*}
1+\theta_{1}z+\theta_{2}z^{2}&=0\\
1-0.3046z-0.3105z^{2}&=0
\end{align*}



## MA Inverse Roots
The `polyroot()` function in R solves for the complex roots of a polynomial.

<grayed>(the coefficients has to be ordered from the constant term to the highest degree term)</grayed>
$$1+\theta_{1}z+\theta_{2}z^{2}=0$$
```{r}
MA2roots <- polyroot(c(1,eq$coef[c("ma1", "ma2")]))
MA2invroots <- 1/MA2roots
print(round(MA2invroots,4))
```
<div style="height: 0.2em;"></div>

So we have <blue>$\zeta_{1}^{-1}=0.7300$</blue> and <blue>$\zeta_{2}^{-1}=-0.4254$</blue>.



## Checking Magnitude

The magnitudes for these inverse roots (takes form $\zeta^{-1}=a+ib$) are:

$$|\zeta^{-1}|=\sqrt{a^{2}+b^{2}} \quad\implies\quad\begin{aligned}|\zeta_{1}^{-1}|&=\sqrt{0.73^{2}+0^{2}}=0.73\;\color{green}{<1}\\
|\zeta_{2}^{-1}|&=\sqrt{(-0.4254)^{2}+0^{2}}=0.4254\;\color{green}{<1}
\end{aligned}$$

The relevant R function is `Mod()`:
```{r}
print(Mod(MA2invroots))
```

<div style="height: 0.2em;"></div>

Since both inverse roots are less than 1 in magnitude <green>$\to$ lies within the unit circle.</green>


## Plot
```{r}
plot(eq)
mtext(paste0("ARMA(",1,",",2,")  "), 
side=3, line=2, cex=1.3, font=2)
```

::: {.absolute top="650px" left="950px" style="font-size: 32px;"}
<blue>
$$\begin{aligned}
&\textsf{Right panel:}\\
&\zeta_{1}^{-1}=0.7300\\
&\zeta_{2}^{-1}=-0.4254
\end{aligned}$$
</blue>
:::

:::


## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}
:::{.callout-question}
Progressively increment the orders of *both* AR and MA models so that we estimate ARMA(1,2), ARMA(2,3), ARMA(3,4) and ARMA(4,5) models. 

<blue>Plot the inverted roots for these four models. Discuss the implication of plots.</blue>
:::
<div style="height: 0.2em;"></div>
```{r eval = FALSE}
for (j in 0:3){
  eq <- Arima(DY, order=c(1+j,0,2+j))
  plot(eq)
  mtext(paste0("ARMA(",1+j,",",2+j,")"), 
    side=3, line=2, cex=1.3, font=2) }
```

:::{.panel-tabset}

## ARMA(1,2)
```{r echo=FALSE}
eq <- Arima(DY, order=c(1,0,2))       # ARMA(1,2)
plot(eq)
mtext(paste0("ARMA(",1,",",2,")"), 
side=3, line=2, cex=1.3, font=2)
```

## ARMA(2,3)
```{r echo=FALSE}
eq <- Arima(DY, order=c(2,0,3))       # ARMA(2,3)
plot(eq)
mtext(paste0("ARMA(",2,",",3,")"), 
side=3, line=2, cex=1.3, font=2)
```

## ARMA(3,4)
```{r echo=FALSE}
eq <- Arima(DY, order=c(3,0,4))       # ARMA(3,4)
plot(eq)
mtext(paste0("ARMA(",3,",",4,")"), 
side=3, line=2, cex=1.3, font=2)
```

## ARMA(4,5)
```{r echo=FALSE}
eq <- Arima(DY, order=c(4,0,5))       # ARMA(4,5)
plot(eq)
mtext(paste0("ARMA(",4,",",5,")"), 
side=3, line=2, cex=1.3, font=2)
```
:::

## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}
:::{.panel-tabset}

## ARMA(1,2)
```{r}
eq <- Arima(DY, order=c(1,0,2)) 
MA2invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2")]))
print(round(MA2invroots,4))         # MA Inverse roots
print(round(Mod(MA2invroots),4))    # Magnitudes of MA inverse roots
```

```{r echo=FALSE}
plot(eq)
mtext(paste0("ARMA(",1,",",2,")"), 
side=3, line=2, cex=1.3, font=2)
```



## ARMA(2,3)
```{r}
eq <- Arima(DY, order=c(2,0,3)) 
MA3invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2", "ma3")]))
print(round(MA3invroots,4))         # MA Inverse roots
print(round(Mod(MA3invroots),4))    # Magnitudes of MA inverse roots
```

```{r echo=FALSE}
plot(eq)
mtext(paste0("ARMA(",2,",",3,")"), 
side=3, line=2, cex=1.3, font=2)
```

::: {.absolute top="470px" left="920px" style="font-size: 30px;"}
Both extra inverted <purple>AR</purple> and <teal>MA</teal> roots are very close to 1 

<blue>$\Rightarrow$ this model for the same time series apparently has suddenly become both non-stationary and non-invertible.</blue>
:::

::: {.absolute top="590px" left="128px" style="font-size: 40px;"}
<purple>$\swarrow$</purple>
:::

::: {.absolute top="590px" left="610px" style="font-size: 40px;"}
<teal>$\swarrow$</teal>
:::

## ARMA(3,4)

```{r}
eq <- Arima(DY, order=c(3,0,4)) 
MA4invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2", "ma3", "ma4")]))
print(round(MA4invroots,4))         # MA inverse roots
print(round(Mod(MA4invroots),4))    # Magnitudes of MA inverse roots
```

```{r echo=FALSE}
plot(eq)
mtext(paste0("ARMA(",3,",",4,")"), 
side=3, line=2, cex=1.3, font=2)
```

::: {.absolute top="470px" left="920px" style="font-size: 30px;"}
With the addition of 2 to both AR and MA orders, we now have <purple>two additional (very close to) non-stationary AR inverse roots</purple> and <teal>two non-invertible MA inverse roots</teal>, all four of which happen to be complex.
:::

::: {.absolute top="525px" left="290px" style="font-size: 40px;"}
<purple>$\nearrow$</purple>
:::

::: {.absolute top="710px" left="290px" style="font-size: 40px;"}
<purple>$\searrow$</purple>
:::

::: {.absolute top="525px" left="770px" style="font-size: 40px;"}
<teal>$\nearrow$</teal>
:::

::: {.absolute top="710px" left="770px" style="font-size: 40px;"}
<teal>$\searrow$</teal>
:::

## ARMA(4,5)
```{r}
eq <- Arima(DY, order=c(4,0,5)) 
MA5invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2", "ma3", "ma4", "ma5")]))
print(round(MA5invroots,4))         # MA inverse roots
print(round(Mod(MA5invroots),4))    # Magnitudes of MA inverse roots
```

```{r echo=FALSE}
plot(eq)
mtext(paste0("ARMA(",4,",",5,")"), 
side=3, line=2, cex=1.3, font=2)
```

::: {.absolute top="560px" left="920px" style="font-size: 30px;"}
Now there are three additional (very close to) <purple>non-stationary AR inverse roots</purple> and <teal>non-invertible MA inverse roots</teal>.
:::

::: {.absolute top="560px" left="290px" style="font-size: 40px;"}
<purple>$\nearrow$</purple>
:::

::: {.absolute top="740px" left="290px" style="font-size: 40px;"}
<purple>$\searrow$</purple>
:::

::: {.absolute top="560px" left="770px" style="font-size: 40px;"}
<teal>$\nearrow$</teal>
:::

::: {.absolute top="740px" left="770px" style="font-size: 40px;"}
<teal>$\searrow$</teal>

:::

::: {.absolute top="620px" left="128px" style="font-size: 40px;"}
<purple>$\swarrow$</purple>
:::

::: {.absolute top="620px" left="610px" style="font-size: 40px;"}
<teal>$\swarrow$</teal>
:::

:::

## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}
:::{.callout-question}
<blue>**What is going on here?**</blue>
:::

:::{.fragment}

:::{.callout-note}
In Tutorial 8, the optimal ARMA model for $\Delta Y_{t}$, chosen by the AICc, is ARMA(1,2). 
Let's assume that it is the correct model.
:::
<div style="height: 0.2em;"></div>
:::

:::{.fragment}
The AIC-selected ARMA(1,2) model can be written symbolically as:

$$(1-\xi_{1}^{-1}L)Z_{t}=(1-\zeta_{1}^{-1}L)(1-\zeta_{2}^{-1}L)U_{t}$$
:::

:::{.fragment}
The AR and MA inverse roots are:
```{r echo=FALSE}
eq <- Arima(DY, order=c(1,0,2)) 
MA2invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2")]))
ARinvroots <- eq$coef[c("ar1")]
mydata <- matrix(c(ARinvroots, MA2invroots), nrow=1)
colnames(mydata) <- c("xi_1", "zeta_1", "zeta_2")
rownames(mydata) <- c("")
print(round(mydata,4))
```

$$(1-0.9038L)Z_{t}=(1-0.73L)(1+0.4254L)U_{t}$$
:::


:::{.fragment}
If ARMA(1,2) is indeed the correct model, then if we increment the specification to ARMA(2,3),
$$(1-\xi_{1}^{-1}L)(1-\color{purple}{\xi_{2}^{-1}}\color{black}{L)Z_{t}=(1-\zeta_{1}^{-1}L)(1-\zeta_{2}^{-1}L)(1\,-}\color{teal}{\,\zeta_{3}^{-1}}\color{black}{L)U_{t}}$$
the “correct” values of the excess terms <purple>$\xi_{2}^{-1}$</purple> and <teal>$\zeta_{3}^{-1}$</teal> should be 0.

:::


## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}
:::{.callout-question}
<blue>**What is going on here?**</blue>
:::
$$(1-0.9038L)Z_{t}=(1-0.73L)(1+0.4254L)U_{t}$$
<div class="horizontal-line"></div>

If ARMA(1,2) is indeed the correct model, then if we increment the specification to ARMA(2,3),
$$(1-\xi_{1}^{-1}L)(1-\color{purple}{\xi_{2}^{-1}}\color{black}{L)Z_{t}=(1-\zeta_{1}^{-1}L)(1-\zeta_{2}^{-1}L)(1\,-}\color{teal}{\,\zeta_{3}^{-1}}\color{black}{L)U_{t}}$$
the “correct” values of the excess terms <purple>$\xi_{2}^{-1}$</purple> and <teal>$\zeta_{3}^{-1}$</teal> should be 0.


:::{.fragment}
<div class="horizontal-line"></div>
<blue>Unfortunately things get more complicated in this specification.</blue>
:::

:::{.fragment}
The AR and MA inverse roots for ARMA(2,3) are:
```{r echo=FALSE}
eq <- Arima(DY, order=c(2,0,3)) 
MA3invroots <- 1/polyroot(c(1,eq$coef[c("ma1", "ma2", "ma3")]))
AR2invroots <- 1/polyroot(c(1,-eq$coef[c("ar1", "ar2")]))
mydata <- matrix(c(AR2invroots, MA3invroots), nrow=1)
colnames(mydata) <- c("xi_1", "xi_2", "zeta_1", "zeta_2", "zeta_3")
rownames(mydata) <- c("")
print(round(mydata,4))
```

<div style="font-size: 80%;">
[$$(1-0.9092L)(1-\color{purple}{(-0.9885)}\color{black}{L)Z_{t}=(1-0.7477L)(1+0.4993L)(1-}\color{teal}{(-0.9564)}\color{black}{L)U_{t}}$$]{.small}
</div>

where <purple>$\xi_{2}^{-1}=-0.9885$</purple> and <teal>$\zeta_{3}^{-1}=-0.9564$</teal> are very much not 0.
:::

## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}
:::{.callout-question}
<blue>**What is going on here?**</blue>
:::

<div style="font-size: 80%;">
$$(1-0.9038L)Z_{t}=(1-0.73L)(1+0.4254L)U_{t}$$

[$$(1-0.9092L)\color{red}{(1+0.9885L)}\color{black}{Z_{t}=(1-0.7477L)(1+0.4993L)}\color{red}{(1+0.9564L)}\color{black}{U_{t}}$$]{.small}
</div>


By introducing the extra terms on both sides we have introduced what is referred to as a <red>“common factor"</red>[, and the <red>"common factor"</red> can be “cancelled” from both sides.]{.fragment}

:::{.fragment}
<div class="horizontal-line"></div>
<blue>What does this mean, and what is the problem?</blue>
:::

:::{.incremental}
- There is only one true data generating process in reality, indicating there is a unique set of AR and MA roots $(\xi_{1}^{-1},\xi_{2}^{-1},\zeta_{1}^{-1},\zeta_{2}^{-1},\zeta^{-1}_{3})$. [Therefore, there should only be a unique set of AR $(\phi_{1},\phi_{2})$ and MA $(\theta_{1},\theta_{2},\theta_{3})$ parameters.]{.fragment}

- Since the common factor can be cancelled out, the inverse roots $\xi_{2}^{-1}$ and $\zeta_{3}^{-1}$ can now take on a range of values **as long as $\xi_{2}^{-1}=\zeta_{3}^{-1}$**. [<red>$\implies$ the model parameters $(\phi_{1},\phi_{2},\theta_{1},\theta_{2},\theta_{3})$ are NOT unique</red>]{.fragment}
[<red>$\implies$ the model is **NOT identified!**</red>]{.fragment}

- This is why the numerical algorithm pushes the excess parameters ($\xi_{2}^{-1}$, $\zeta_{3}^{-1}$) towards the unit circle, even though we might expect them to shrink towards zero.
:::


## Part A - Exercise 1: Common factors in ARMA models {auto-animate="true"}

<blue>**Some practical observations:**</blue>

:::{.incremental}
- <blue>If an ARMA($p,q$) model has inverse roots that nearly cancel, especially near the unit circle, you should reconsider whether smaller orders would suffice.</blue>

- <blue>As AIC/AICc is balancing between model complexity and fit quality, it would usually be the case that it will work to avoid this issue.</blue>

  - However, it is worth acknowledging that AICc is not perfect and might select   models that is too large.


- <blue>When doing model selection searches, it is generally advised to avoid increasing the order of both AR and MA components simultaneously.</blue>

  - It is better to increase AR or MA components separately, in which case this potential identification problem should be avoided.
  
  - Brute-force AIC search over ($p,q$) works for small grids but can cause instability at larger orders <teal>$\to$ there exists smarter methods (e.g. `auto.arima`) avoid this.</teal>
  
- <blue>More pragmatically, at least in economics and finance it has become more common to rely simply on AR models rather than ARMA.</blue>

  - Avoids the common factor problem, quicker (can use OLS), more stable than ARMA, and simplifies extensions to multivariate models (e.g. VAR).
  
:::

## Part B: Multiplicative Models: Seasonal Autoregression {auto-animate="true"}


:::{.fragment}
An alternative model for seasonality is the seasonal AR model. <grayed>(There is also a seasonal MA model, which follows the same principles.)</grayed>
:::

:::{.fragment}
<div class="horizontal-line"></div>
A simple example is the <blue>first order seasonal AR (SAR) model</blue>, specified here for quarterly data:

$$
Z_{t}=\Phi_{1}Z_{t-4}+U_{t}\quad \iff\quad (1-\Phi_{1}L^{4})Z_{t}=U_{t}
$$

That is, $Y_{t}$ is modeled directly in terms of the observation from one year ago. This is four time periods for quarterly data, hence the $Z_{t-4}$ term.
:::

:::{.fragment}
<div class="horizontal-line"></div>
In general we can have higher order seasonal AR models:
$$
(1-\Phi_{1}L^{4}-\cdots-\Phi_{P}L^{4P})Z_{t}=U_{t}
$$

which expresses the forecasts for $Y_{t}$ in terms of observations from 
$1,\dots,P$ years ago.
:::

## Part B: Multiplicative Models: Seasonal Autoregression {auto-animate="true"}

Most commonly the SAR model is combined with an AR model, so that seasonal and non-seasonal components are permitted. 


:::{.fragment}
<div class="horizontal-line"></div>
This is done in a multiplicative manner using lag polynomials. 

For example an AR(1),SAR(1) model is
$$(1-\phi_{1}L)(1-\Phi_{1}L^{4})Z_{t}=U_{t}$$
This is essentially a restricted AR(5) model. <blue>$\to$ Exercise 2(a)</blue>

:::

:::{.fragment}
<div class="horizontal-line"></div>
In general an AR($p$),SAR($P$) model is
$$(1-\phi_{1}L-\cdots-\phi_{p}L^{p})(1-\Phi_{1}L^{4}-\cdots-\Phi_{P}L^{4P})Z_{t}=U_{t}$$
The selection of $p$ and $P$ can be handled in the usual way using AIC and residual autcorrelation checks with Ljung-Box testing.
:::

## Exercise 2(a)  {auto-animate="true"}

:::{.callout-question}
Expand the lag polynomials in the AR(1),SAR(1) model to show that it is equivalent to an AR(5) model with particular parameter restrictions.
$$(1-\phi_{1}L)(1-\Phi_{1}L^{4})Z_{t}=U_{t}$$
:::

<div style="height: 0.5em;"></div>
:::{.fragment}
We can algebraically manipulate lag polynomials like any other polynomial:

\begin{align}
(1-\phi_{1}L)(1-\Phi_{1}L^{4})&=1-\phi_{1}-\Phi_{1}L^{4}+\phi_{1}\Phi_{1}L\cdot L^{4}\\
&=1-\phi_{1}-\Phi_{1}L^{4}+\phi_{1}\Phi_{1}L^{5}
\end{align}

:::

:::{.fragment}
Thus the model can be re-written as:

$$(1-\phi_{1}L-\Phi_{1}L^{4}+\phi_{1}\Phi_{1}L^{5})Z_{t}=U_{t}$$
:::

:::{.fragment}
$$\implies\qquad Z_{t}=\phi_{1}Z_{t-1}+\Phi_{1}Z_{t-4}-\phi_{1}\Phi_{1}Z_{t-5}+U_{t}$$
:::

:::{.fragment}
Comparing this model with the standard AR(5) setup:
$$Z_{t}=\phi_{1}^* Z_{t-1}+\phi_{2}^* Z_{t-2}+\phi_{3}^* Z_{t-3}+\phi_{4}^* Z_{t-4}+\phi_{5}^* Z_{t-5}+U_{t}$$
:::

:::{.fragment}
the restrictions are $\phi_{1}^*=\phi_{1}$, $\phi_{2}=\phi_{3}=0$, $\phi_{4}^{*}=\Phi_{1}$ and $\phi_{5}^*=-\phi_{1}\Phi_{1}$.
:::
<div style="height: 0.5em;"></div>


## Exercise 2(b) {auto-animate="true"}

:::{.callout-question}
Using `RetailSales.csv` from Week 1/2, carry out a model selection search over AR($p$), SAR($P$) models with maximum of 4 for both $p$ and $P$.

$$(1-\phi_{1}L-\cdots-\phi_{p}L^{p})(1-\Phi_{1}L^{4}-\cdots-\Phi_{P}L^{4P})Z_{t}=U_{t}$$
What is the longest lag of $Z_{t}$ implied by this AR,SAR model?

:::

<div style="height: 0.5em;"></div>
<blue>**Initialisation:** Copy the following to set up in the usual way</blue>

```{r}
library(forecast)

# Set up log retail sales, 2000q1 to 2019q4
dt <- read.csv("RetailSales.csv")
Retail_m <- ts(dt$Total, frequency=12, start=c(1982,4), end=c(2024,9))
Retail_q <- aggregate(Retail_m, nfrequency=4)
log_Retail_q <- log(Retail_q)
Y <- window(log_Retail_q, start=c(2000,1), end=c(2019,4))

# Linear trend
Time <- time(Y)

```
<div style="height: 0.5em;"></div>
<grayed>(continued next slide...)</grayed>

## Exercise 2(b)  {auto-animate="true"}

Set up deterministics:
```{r}
# GFC trend break
gfc <- 2008.5
Time_postgfc <- 1*(Time>gfc)*(Time-gfc)

# Quarterly dummies
QD <- seasonaldummy(Y)

# Deterministics without seasonal dummies
X <- cbind(Time, Time_postgfc)

# Deterministics with seasonal dummies
XQD <- cbind(Time, Time_postgfc, QD)
```
<div style="height: 0.5em;"></div>


Define samples for estimation and forecast evaluation:
```{r}
# Estimation sample
t_est <- which(Time<2019)
Y <- window(log_Retail_q, start=c(2000,1), end=c(2018,4))

# Forecast evaluation sample
t_for <- which(Time>=2019)
Yf <- window(log_Retail_q, start=c(2019,1), end=c(2019,4))
```


## Exercise 2(b)  {auto-animate="true"}

:::{.callout-question}
Using `RetailSales.csv` from Week 1/2, carry out a model selection search over AR($p$), SAR($P$) models with maximum of 4 for both $p$ and $P$.

$$(1-\phi_{1}L-\cdots-\phi_{p}L^{p})(1-\Phi_{1}L^{4}-\cdots-\Phi_{P}L^{4P})Z_{t}=U_{t}$$
What is the longest lag of $Z_{t}$ implied by this AR,SAR model?

:::
<div style="height: 0.2em;"></div>
```{r}
pmax <- 4
Pmax <- 4
AICc <- matrix(nrow=1+pmax, ncol=1+Pmax)
rownames(AICc) <- paste0("AR",0:pmax)
colnames(AICc) <- paste0("SAR",0:Pmax)
LBp <- AICc

for (p in 0:pmax){
  for (P in 0:Pmax){

    # Specification of AR(p), SAR(P) model:
    eq_sar <- Arima(Y, order=c(p,0,0), xreg=X[t_est,],
            seasonal=list(order=c(P,0,0), period=4))

    AICc[1+p,1+P] <- eq_sar$aicc
    LBp[1+p,1+P] <- Box.test(eq_sar$residuals, type="Ljung-Box", 
                                fitdf=p+P, lag=12)$p.value
  }
}
```
<div style="height: 0.5em;"></div>


## Exercise 2(b)  {auto-animate="true"}
 
:::{.callout-question}
Using `RetailSales.csv` from Week 1/2, carry out a model selection search over AR($p$), SAR($P$) models with maximum of 4 for both $p$ and $P$.

$$(1-\phi_{1}L-\cdots-\phi_{p}L^{p})(1-\Phi_{1}L^{4}-\cdots-\Phi_{P}L^{4P})Z_{t}=U_{t}$$
What is the longest lag of $Z_{t}$ implied by this AR,SAR model?

:::

<div style="height: 0.5em;"></div>
**Results:**

:::{.panel-tabset}

## AICc values
```{r}
print(round(AICc, 4))
```

## Model selection search
```{r}
print(AICc==min(AICc))
```
<div style="height: 0.5em;"></div>

<blue>The selected model has $p=1$ and $P=3$.</blue> $\implies$ longest lag of $Z_{t}$ is 13.

$$\color{teal}{\boxed{\textsf{\small\;found by expanding this }\to\;}}\quad\color{black}{ (1-\phi_{1}L)(1-\Phi_{1}L^{4}-\Phi_{2}L^{8}-\Phi_{3}L^{12})Z_{t}=U_{t}}$$

## Residual autocorrelation

:::: {.columns}

::: {.column width="50%"}
```{r}
print(round(LBp,3))
```
:::

::: {.column width="50%"}
```{r}
print(1*(LBp<0.05))
```
:::

::::


<div style="height: 0.5em;"></div>

Since $p=0.739>0.05$, we can maintain $H_0$ of no residual autocorrelation

<green>$\implies$ the AR(1),SAR(3) model is not misspecified.</green>
:::


## Exercise 2(b)  {auto-animate="true"}
:::{.callout-question}
Use the a model with the selected $p$ and $q$ to forecast the four quarters of 2019 (i.e. 1,2,3,4-steps ahead), and compute their RMSE.
:::
<div style="height: 0.5em;"></div>
```{r}
p <- 1
P <- 3
eq_sar <- Arima(Y, order=c(p,0,0), xreg=X[t_est,],
        seasonal=list(order=c(P,0,0), period=4))

# Forecast through 2019
Yf_sar <- forecast(eq_sar, xreg=X[t_for,])$mean

# Compute RMSE
RMSE_sar <- sqrt(mean((Yf-Yf_sar)^2))
```
<div style="height: 0.5em;"></div>

The forecasts from AR(1),SAR(3) are:
```{r}
print(Yf_sar)
```
<div style="height: 0.5em;"></div>

<blue>The RMSE for AR(1),SAR(3) forecasts is:</blue>
```{r}
print(RMSE_sar)
```


## Exercise 2(c) {auto-animate="true"}

:::{.callout-question}
Previously we have identifed an AR(3) model is suitable when also using a linear trend with GFC break and quarterly dummies. Estimate this model, obtain the forecasts for the four quarters of 2019, and compute their RMSE.
:::
<div style="height: 0.5em;"></div>

```{r}
# Estimation of AR(3) with XQD as deterministic matrix (with seasonal dummies)
eq_qd <- Arima(Y, order=c(3,0,0), xreg=XQD[t_est,])

# Forecast through 2019
Yf_qd <- forecast(eq_qd, xreg=XQD[t_for,])$mean

# Compute RMSE
RMSE_qd <- sqrt(mean((Yf-Yf_qd)^2))
```
<div style="height: 0.5em;"></div>

The forecasts from AR(3) are:
```{r}
print(Yf_qd)
```
<div style="height: 0.5em;"></div>

<blue>The RMSE for AR(3) forecasts is:</blue>
```{r}
print(RMSE_qd)
```
::: {.absolute top="850px" left="290px" style="font-size: 30px;"}
<purple>$\longleftarrow$ The SAR model has slightly lower RMSE (at 0.0078) than the AR(3) with quarterly dummies.</purple>
:::

## Exercise 2(d) {auto-animate="true"}

:::{.callout-question}
Repeat the search in (b) but with quarterly dummies also included. This is combining both quarterly dummies and seasonal AR modelling to capture the seasonality in the data.
:::

<div style="height: 0.5em;"></div>
:::{.fragment}

<blue>First, observe the AICc values:</blue>

```{r}
# AICc for AR(3) with seasonal dummies
print(eq_qd$aicc)
```
<div style="height: 0.5em;"></div>
```{r}
# AICc for Seasonal AR: AR(1),SAR(3)
print(eq_sar$aicc)
```
<div style="height: 0.5em;"></div>
The model with quarterly dummies is clearly preferred by the AICc to the seasonal autoregression.

:::

## Exercise 2(d) {auto-animate="true"}

:::{.callout-question}
Repeat the search in (b) but with quarterly dummies also included. This is combining both quarterly dummies and seasonal AR modelling to capture the seasonality in the data.
:::

<div style="height: 0.5em;"></div>

We would like to check if including the quarterly dummies in the seasonal AR model may improve the model further: <grayed>(conduct the grid search again, but use `XQD` instead)</grayed>

<div style="height: 0.6em;"></div>
```{r}
pmax <- 4
Pmax <- 4
AICc <- matrix(nrow=1+pmax, ncol=1+Pmax)
rownames(AICc) <- paste0("AR",0:pmax)
colnames(AICc) <- paste0("SAR",0:Pmax)
LBp <- AICc

for (p in 0:pmax){
  for (P in 0:Pmax){
    eq_sar <- Arima(Y, order=c(p,0,0), xreg=XQD[t_est,],  # <-- changed
          seasonal=list(order=c(P,0,0), period=4))
    AICc[1+p,1+P] <- eq_sar$aicc
  }
}
```


## Exercise 2(d) {auto-animate="true"}

:::{.callout-question}
Repeat the search in (b) but with quarterly dummies also included. This is combining both quarterly dummies and seasonal AR modelling to capture the seasonality in the data.
:::

<div style="height: 0.5em;"></div>

We would like to check if including the quarterly dummies in the seasonal AR model may improve the model further: <grayed>(conduct the grid search again, but use `XQD` instead)</grayed>

<div class="horizontal-line"></div>

Thus when quarterly dummies are used...

```{r}
print(AICc==min(AICc))
```

<div style="height: 0.5em;"></div>
:::{.fragment}
... AR(3) remains the preferred specification [<teal>$\to$ there is no further improvement in model fit by also using seasonal autoregressive components.</teal>]{.fragment}
:::

## {data-background-color="#000F46"}
<div style="height: 85vh; display: flex; align-items: center; justify-content: center; color: white; font-size: 2em;">
Thank you for coming!
</div>



<script>
document.querySelectorAll("myimage").forEach(el => {
  const src = el.getAttribute("src");
  const width = el.getAttribute("width") || "100%";

  const img = document.createElement("img");
  img.src = src;
  img.style.width = width;
  img.style.maxWidth = "100%";  // prevent overflow
  el.style.display = "block";
  el.style.textAlign = "center";
  
  el.appendChild(img);
});
</script>